Loan word maps

### 2025-03-12 analysis/paper/paper.qmd
added data and map production code

```{r}
#| label: get-data
#| eval: false
# Read data
langloc <- readr::read_csv(here::here('analysis/data/raw_data/language_location.csv')) |>

@@

```{r}
# read classified loan word data

# Define the file path
file_path <- here::here('analysis/data/raw_data/cleaned exports.xlsx')

# Get the names of all the sheets in the Excel file
sheet_names <- readxl::excel_sheets(file_path)

# Read each sheet and store it as a separate data frame
for (sheet in sheet_names) {
  assign(sheet, readxl::read_excel(file_path, sheet = sheet, col_names = TRUE)) 
}

add the location data

```{r}
# add location to the gold sheet

# merge by glottolog code 
gold <- gold |> 
  #join by glottolog code
  dplyr::left_join(langloc, by = c("lang" = "GlottCode"))

# split off where glottolog code matching was successful
gold_a <-  gold |> dplyr::filter(!is.na(LanName))

gold_a <- gold |> filter(!is.na(LanName))

#where the glottolog code is not available, merge by Sahelia name
gold <- gold |> 
  dplyr::mutate(id2 = ifelse(is.na(LanName), id2, NA))

# Perform the first join on the primary id1
merged_data <- data1 %>%
  left_join(data2, by = c("id1" = "id"))

>>>>

```{r}
#| label: get-data
#| eval: false
# Read data
langloc <- readr::read_csv(here::here('analysis/data/raw_data/glottologAfricanLanguages.csv')) |>

  #retain only relevant variables
  dplyr::select(glott_code, latitude, longitude, LanName) |>

@@

```{r}
# read classified loan word data
# Create a vector of file paths
objs <- here::here("analysis", "data", "raw_data", c("gold.csv", 
                                                     "fonio.csv",
                                                     "millet.csv",
                                                     "salt.csv",
                                                     "iron.csv"))

# read the csv files into a list of data frames
data_list <- purrr::map(objs, readr::read_csv)

# Add the location data to the word lists

# Function to join location data to each data frame
join_geometry <- function(df) {
  df |> 
    dplyr::left_join(langloc, by = "glott_code") |> 
    sf::st_as_sf()  # Convert the result to an sf object if needed
}

# Apply the join function to each data frame in the list
data_list <- purrr::map(data_list, join_geometry)
```

Function to make interactive leaflet maps
```{r}
create_leaflet_map <- function(data_sf, form_column = "form", background = "colour") {

  # Ensure the input data is an sf object and contains the 'form' column
  if (!inherits(data_sf, "sf")) {
    stop("The input data must be an sf object.")
  }
  if (!form_column %in% colnames(data_sf)) {
    stop(paste("The column", form_column, "is not found in the data."))
  }

  # Filter out rows with NA values in the "form" column
  data_sf <- data_sf |>
    dplyr::filter(!is.na(!!rlang::sym(form_column)))

  # Set the tile provider based on the 'background' argument
  tile_provider <- switch(
    background,
    "colour" = leaflet::providers$Esri.WorldPhysical,  # Default color tile
    "bw" = leaflet::providers$Esri.WorldGrayCanvas,    # Black-and-white tile
    stop("Invalid value for 'background'. Choose 'colour' or 'bw'.")
  )

  # Define a color palette for the "form" column (using Set3 palette)
  color_pal <- leaflet::colorFactor(palette = "Set3", domain = data_sf[[form_column]])

  # Add a new column with the corresponding colors based on the 'form' column
  data_sf <- data_sf |>
    dplyr::mutate(color = color_pal(!!rlang::sym(form_column)))  # Dynamically reference the column

  # Adjust overlapping points by slightly jittering their coordinates
  # Initially apply fixed jitter for the map creation
  jitter_scale <- 0.1000  # Set the base jitter scale factor

  data_sf <- data_sf |>
    dplyr::mutate(
      # Apply a small random jitter to coordinates for initial display
      jittered_lon = sf::st_coordinates(geometry)[, 1] + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale),
      jittered_lat = sf::st_coordinates(geometry)[, 2] + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale)
    )

  # Create the leaflet map
  leaflet_map <- leaflet::leaflet(data_sf) |>
    leaflet::addProviderTiles(tile_provider) |>
    leaflet::addLabelOnlyMarkers(
      lng = ~jittered_lon,  # Use jittered longitude
      lat = ~jittered_lat,  # Use jittered latitude
      label = purrr::map(
        glue::glue("<span style='color:{data_sf$color}; font-weight: bold; font-size: 12px;'>{as.character(data_sf[[form_column]])}</span>"),
        htmltools::HTML
      ),
      labelOptions = leaflet::labelOptions(
        noHide = TRUE,
        direction = 'top',
        textOnly = TRUE
      )
    )

  # Add JavaScript to dynamically adjust jitter based on zoom level
  leaflet_map <- leaflet_map |>
    htmlwidgets::onRender(
      "
      function(el, map) {
        // Function to apply jitter based on zoom level
        function applyJitter() {
          var zoomLevel = map.getZoom();
          var jitterScale = 0.1 * Math.pow(1.7, zoomLevel); // Adjust jitter with zoom level

          // Adjust the jitter of each marker
          map.eachLayer(function(layer) {
            if (layer instanceof L.Marker) {
              var latLng = layer.getLatLng();
              var newLat = latLng.lat + (Math.random() - 0.5) * jitterScale;
              var newLng = latLng.lng + (Math.random() - 0.5) * jitterScale;
              layer.setLatLng([newLat, newLng]);
            }
          });
        }

        // Apply jitter on map zoom change
        map.on('zoomend', function() {
          applyJitter();
        });

        // Apply jitter initially
        applyJitter();
      }
      "
    )

  # Return the map object
  return(leaflet_map)
}
```

Function to save a version of every map in the list of data
```{r}
# Function to create and save leaflet maps for all data frames in data_list
save_maps <- function(data_list, form_column = "form", background = "colour") {

  # Use purrr::map2 to iterate over the list of data frames and generate maps
  purrr::map2(data_list, seq_along(data_list), ~ {
    # Create the map using the create_leaflet_map function
    map <- create_leaflet_map(.x, form_column, background)

    # Generate the filename based on the index or data frame name
    filename <- paste0("map_", .y, ".html")

    # Save the map as an HTML file in the 'analysis/figures' folder
    htmlwidgets::saveWidget(map, here::here("analysis/figures", filename))
  })
}
```

Save an html version of all objects
```{r}
save_maps(data_list, background = "bw")
```
